TTS2016R: A dataset to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area, Canada

### 2022-10-23 Manuscript-Data-Package/Manuscript-Data-Package.tex
submitted this version to EPB
draft of revised manuscript

This manuscript presents the open data product
\href{https://github.com/soukhova/TTS2016R}{\{TTS2016R\}}. Open data
products are the result of turning source data (open or otherwise) into
accessible information that adds value to the original inputs
\citep[see][]{Arribas2021open}. The product presented in this paper is
an \texttt{R} data package which currently consists of three objects
which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or
are curated to facilitate the use and analysis of TTS data. This package
includes person-to-jobs origin-destinations, traffic analysis zone (TAZ)
boundaries, and planning/municipality boundaries for the Greater Golden
Horse area (GGH) located in southern Ontario, Canada
\citep{data_management_group_tts_2018}. In addition, the package
includes TAZ centroid-to-centroid travel times by car computed using
package \{r5r\} \citep{Pereira2021r5r}. The aim of this paper is to walk
readers through the empirical home-based work commute data set,
illustrate the calculation of an impedance function that can be used to
calculate accessibility to employment, and invite its use in other
applications.

@@

\{TTS2016R\} includes counts of fully-employed population by place of
residence (origin) and counts of full-time jobs by place of work
(destination) aggregated by TAZ (n=3,764 within the survey boundaries).
TAZ typically are defined based on land-use and population demographics
in order to estimate the number of trips produced and attracted to each
zone \citep{meyer_urban_2001}. As such, each TAZ is uniquely identified
using the GTA06 Zoning System which can be used to join to the
origin-destination table (i.e., trips taken).

The number of jobs (3,081,885) and workers (3,446,957) in this package
are organized in the form of an origin-destination table which is
indicative of home-to-work commute patterns (there are 3,446,957
potential interactions). These data were retrieved from the TTS Data
Retrieval System on October 28, 2021 and reflect the potential
interaction of full-time employed people and jobs within the GGH survey
boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by
the 2016 TTS methodology \citep{data_management_group_tts_2018}.

Also included in \{TTS2016R\} are travel times and cost of travel from
origin to destination by car; travel times are calculated using the
\texttt{R} package \{r5r\}. These travel times are useful to estimate
the cost of travel and to calculate impedance functions, among other
possible applications. For simplicity, all interactions within
\{TTS2016R\} are assumed to be taken by car, and the travel time is
calculated from an origin TAZ centroid to a destination TAZ centroid.
The centroid is snapped to the nearest street line by \texttt{r5r} and
the travel time is calculated for all trips assuming a car travel mode.
Additionally, only travel times less than or equal to 180 mins (3 hrs)
are calculated; this threshold represents 99\% of trip's travel times
which are summarized in Figure \ref{fig:TTS-16-survey-area}.

@@

\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ. Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).}\label{fig:TTS-16-survey-area}
\end{figure}

\hypertarget{employed-individuals-and-jobs}{%
\subsection{Employed individuals and
jobs}\label{employed-individuals-and-jobs}}

The origin-destination table (i.e., trips) consists of a
cross-tabulation of people who are employed full-time by place of GGH
residence (origin) and places of GGH employment (destination) using the
GTA06 TAZ zoning system. It is important to note that the total number
of workers and jobs is the TTS 2016 region are not equal but the number
of trips taken are equal to the number of workers. Since the outer
boundaries of the TTS are permeable, workers who reside within the
boundaries but travel outside of the boundaries are counted as workers
within an origin TAZ, while jobs in TAZ that are filled by workers who
reside outside the GGH boundaries are \emph{unknown} since they were not
surveyed. This mismatch results in the total number of workers being
1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
3,081,885 jobs). As such, the origin-destination table contained in
\{TTS2016R\} offers a perspective on all workers in the GGH and their
home-based trips to places of GGH employment.

@@

Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers
and jobs per TAZ. It can be observed that the spatial distribution of
jobs and workers is unequal, which is indicative of a jobs-housing
imbalance that can impact accessibility in a region
\citep{Levine1998rethinking}. It can also be seen that there is a higher
number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
is higher than the mean of jobs (i.e., 916 workers : 819 jobs) the
number TAZ with an extreme number of jobs at the highest and lowest
percentiles is significantly higher than the number of workers.

@@

As mentioned, \{TTS2016R\} also includes travel time data for each
home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
This travel time corresponds to a car commute calculated using the R
package \{r5r\} (see descriptive statistics in Figure
\ref{fig:TTS-16-survey-area}). It is important to note that travel times
within this data set are calculated assuming only car travel and one
departure time for all origins. These assumptions are not completely
realistic since we know a small proportion of trips are taken by non-car
modes and travel time departures varies. However, it is not possible
from the data retrieval system to obtain higher order tabulations so we
carry on with the assume that all trips are taken by one-time departure
car trip.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.}\label{fig:plot-tt-ttpertrip}
\end{figure}

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total
travel time resembles the spatial trend distribution in the number of
employed people in the previous plot (Figure
\ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the
average travel time is distinct from other plots presented so far. We
can see that in areas around the south-eastern border that make up the
Greater Toronto and Hamilton Area (GTHA), Niagara and Waterloo, the
average travel times are moderately low. Further from these areas,
travel times are higher. Interestingly, even in eastern areas (e.g.,
Peterborough) with high employment and high job concentration, average
travel time is higher than within the GTHA.

@@

\begin{equation}
\label{gamma-dist}
\begin{array}{l}\ 
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
\end{array}
\end{equation}

@@

The open data product introduced in this paper shares tables for
home-to-work related data from the 2016 TTS. In addition, inter-centroid
travel time tables are calculated, and the planning/municipality
boundaries are included as a compliment. This open data product,
\{TTS2016R\}, is freely available to explore in an \texttt{R}
environment. One possible use of this data, as showcased in this paper,
is the calibration of impedance functions which in turn can be used for
accessibility analysis.

>>>>

This manuscript presents the open data product
\href{https://github.com/soukhova/TTS2016R}{\{TTS2016R\}}. Open data
products are the result of turning source data (open or otherwise) into
accessible information that adds value to the original inputs
\citep[see][]{Arribas2021open}. The product presented in this paper is a
\texttt{R} data package which currently consists of a fusion of objects
from a variety of sources: home-to-work flows and trips by mode are
sourced from the 2016 Transportation Tomorrow Survey (TTS)
\citep{data_management_group_tts_2018}, estimated travel times are
calculated, and boundary files from the TTS and from the Canadian
Census.

What is a \texttt{R} data package? A \texttt{R} data package contains
code, data, and documentation in a standardised collection format that
can be installed by \texttt{R} users through a centralized software
repository such as CRAN (the Comprehensive R Archive Network) and
GitHub. \{TTS2016R\} is freely available on GitHub for all to install
and freely use in the spirit of open and reproducible research.
Currently, \{TTS2016R\} includes full-time home-based work-to-job origin
destinations counts and mode-specific trip numbers retrieved from the
2016 TTS, traffic analysis zone (TAZ) boundaries, and municipality,
planning, and census metropolitan area boundaries for the Greater Golden
Horse area (GGH) located in southern Ontario, Canada. In addition, the
package includes TAZ centroid-to-centroid travel times by car, transit,
cycling, and walking mode computed using package \{r5r\}
\citep{Pereira2021r5r}.

The aim of this paper is to walk readers through the data sets,
illustrate a use case (i.e., the calculation of an impedance function
that can be used to calculate accessibility to employment), and invite
others to experiment in its uses and applications. Though data from the
TTS is freely available to the public through the
\href{https://dmg.utoronto.ca/idrs/index}{TTS Data Retrieval System},
the raw data can be technically demanding, cumbersome to work with, and
requires multiple software to process. By pre-processing the data in a
\texttt{R} environment and packaging it alongside complimentary data,
\{TTS2016R\} offers a slice of the TTS data which can be immediately
used by \texttt{R} users to analysis patterns of commuting to work in
the region. Anticipate this package to grow in the future: it currently
provides an open infrastructure for additional TTS or complimentary data
sets to be amended by the authors and the open-source community in the
future by request.

@@

Currently, \{TTS2016R\} includes counts of full-time employed population
by place of residence (origin), counts of full-time usual place of work
(destination), number of trips to work by mode, and the calculated
potential travel time of the trips in the GGH. The GGH (and hence the
TTS survey area) is displayed in Figure \ref{fig:TTS-16-survey-area}.

This data is aggregated and available at the level of TAZ: TAZ are a
spatial unit of analysis typically used to estimate the number of trips
produced and attracted to each zone \citep{meyer_urban_2001}. They are
thus defined by transportation planners for a region based on
intra-similarity and inter-dissimilarity between land-use and population
demographics. Within the GGH boundaries, 3,764 TAZ are specified and
each TAZ is uniquely identified using the GTA06 Zoning System: the
survey boundary is discussed in the 2016 TTS methodology and defined by
the TTS \citep{data_management_group_tts_2018}. The TAZ range between
\(\ge\)\texttt{0.019} km\^{}2 in spatial area to a maximum of 879
km\^{}2 (median: 1.3 km\^{}2 and 3rd quantile: 2.8 km\^{}2).

\begin{figure}[H]

@@

\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, OD trips by origin TAZ, and calculated OD car travel time (TT) per origin TAZ. 3,507 trips were not assigned travel times as they are longer than 180 mins. Spatial boundary files are retrieved from the TTS which define the survey area: the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.}\label{fig:TTS-16-survey-area}
\end{figure}

\hypertarget{full-time-employed-people-and-associated-places-of-employment}{%
\subsection{Full-time employed people and associated places of
employment}\label{full-time-employed-people-and-associated-places-of-employment}}

In the GGH, there are 3,446,957 workers, 3,081,900 jobs, and 3,282,611
work-related trips (for the 2016 TTS survey day). The values are
organized within the origin destination (OD) table in the \{TTS2016R\}
package and are derived from the cross-tabulation by person and by trip
for the full-time employed population and associated places of
employment. The TTS is a proportionally representative survey, hence the
values included in \{TTS2016R\} are adjusted to reflect the GGH
population.

It is important to note that the total number of full-time workers and
jobs in the TTS 2016 region are not equal. Since the outer boundaries of
the TTS are permeable, workers who reside within the boundaries but have
workplaces that are outside of the boundaries are counted as workers
within an origin TAZ, while jobs in TAZ that are filled by workers who
reside outside the GGH boundaries are \emph{unknown} since they were not
surveyed. This mismatch results in the total number of workers being
1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
3,081,900 jobs). As such, the OD table contained in \{TTS2016R\} offers
a perspective on all workers in the GGH and their home-based trips to
places of GGH employment.

The count of links and trips made by the full-time working population
and associated full-time place of employment per unique OD pair are
quite variable. TAZ contain between 0 to 8,491 workers (median: 464, 3rd
quantile: 1,378), 0 to 41,821 jobs (median: 244, 3rd quantile: 700), and
generate between 0 to 241 trips (median: 15, 3rd quantile: 42).

@@

Figure \ref{fig:tts-workers-jobs-plot} presents the number of employed
people and associated jobs per TAZ. It can be observed that the spatial
distribution of jobs and workers is unequal, which is indicative of a
jobs -housing imbalance that can impact accessibility in a region
\citep{Levine1998rethinking}. It can also be seen that there is a higher
number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
is higher than the mean of jobs (i.e., 916 workers : 819 jobs) the
number TAZ with an extreme number of jobs at the highest and lowest
percentiles is significantly higher than the number of workers.

@@

Also included in \{TTS2016R\} are the estimated travel times between OD
as summarized in descriptive statistics table in Figure
\ref{fig:plot-tt-ttpertrip}; travel times are calculated using the
package \{r5r\}. \{r5r\} interfaces with the java-based \{r5\} routing
engine developed separately by \href{https://conveyal.com/}{Conveyal}.
The inputs to \{r5r\} for this data package were the desired mode, a
maximum travel time threshold of 180 minutes, the geo-coded origin
destination pairs based on the centroids of the TAZ, and the static Open
Street Map road network of Ontario. A travel time threshold of 180
minutes was selected since it captures almost all potential OD
interactions.

Additionally, the car mode was included since it is a critically
important commute mode in the GGH. 2,598,379 of the trips are made using
a car mode out of the total 3,282,611 work-related trips according to
the TTS 2016 data (i.e., 79\% of trips are taken by car).

These travel times are useful addition to \{TTS2016R\} since they are
not included in the TTS Data Retrieval System but they are vitally
important to estimate the cost of travel and associated impedance
functions, among other possible applications. If the readership is
interested in additional information regarding the travel time
computation, please see the calculation notebook in the documentation of
\{TTS2016R\} and details about \{r5r\} at the
\href{https://ipeagit.github.io/r5r/index.html}{r5r package website}.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS. Planning boundaries of Niagara, Waterloo, and the Toronto census metropolitan area are drawn with purple, brown and blue borders, respectively, and discussed. }\label{fig:plot-tt-ttpertrip}
\end{figure}
\newpage

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total
travel time resembles the spatial trend distribution in the number of
employed people in the previous plot (Figure
\ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the
average travel time is distinct from other plots presented so far. For
instance, we can see that in areas around the south-eastern border such
as Niagara and Waterloo (purple and brown borders), the average travel
times are moderately low. Additionally, travel times (by car) within the
core of the Toronto census metropolitan area (CMA) (blue) is also
moderately since traffic congestion is not reflected in the travel time
estimations. Further from these areas, travel times are higher.

@@

\begin{equation}
\label{gamma-dist}
\begin{array}{l}\ 
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
\end{array}
\end{equation}\textbackslash end\{equation\}

@@

\{TTS2016R\}, the open data package introduced in this paper fuses
multiple sources of data. It includes an OD cross-tabulation by person
and by trip mode table for home-to-work commute data from the 2016 TTS
alongside complimentary boundaries and estimated travel times. The value
of this data package is in its transparency, easy of access, and its
open infrastructure for the addition of complimentary data sets in the
future. Using \texttt{R} users can immediately and easily explore GGH
commute flow trends as well as suggestion further amendments to the
package by request. One possible use of this data, as showcased in this
paper, is the calibration of impedance functions which in turn can be
used for accessibility analysis.
